### 2020-09-18 latex/vnhcs.tex
First full edit of VNHCs chapter

\section{Virtual Nonholonomic Constraints}
% Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing, who wants to go as high as possible. The
child will push off the ground and start swinging with small oscillations,
then extend and retract their feet appropriately to gain energy. 
If a roboticist were designing a machine to do this task, they might design a
control mechanism which makes the legs track a trajectory over time. For
example, they might tell the robot to extend and retract its legs every two
seconds. In ideal situtations, this technique would work perfectly because the
leg motion is synchronized with the swing to gain energy as fast as possible.

% TODO: Get some sources here. Show with research that trajectory control is
% unsafe, and that path tracking is better.
Most children have an adult pushing the swing to help them go higher, or perhaps
they are swinging on a windy day. In either case, they adjust their leg
motion accordingly when presented with these external disturbances, without
keeping track of the time it takes to kick their legs. The
standard control technique of tracking a function of time, known as
\textit{trajectory tracking}, does not work as well as human motion 
because the disturbance affecting the swing will desynchronize the leg motion
and the swing, thereby stopping the energy-gaining effects. 
As has been observed in the field of mobile robotics,
a trajectory tracking controller which has lagged behind due to a disturbance
will do whatever is required to ``catch up" to
the trajectory. This can cause major safety issues, which is why mobile
roboticists prefer to design controllers which drive a vehicle towards a path in
space rather than a trajectory in time.

This path tracking approach has been extended to biologically-inspired robotics
in a method known as virtual holonomic constraints (VHCs).
Instead of a robot's actuators tracking
a trajectory over time, VHCs use the actuators to enforce a relation
\(h(q) = 0\) of the configuration \cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like swimming robot
\cite{vhc_snake}.

The downside to VHCs is that they do not use the additional information imparted
by the generalized velocities of a system.
For the child on a swing, whether they extend or retract their legs
depends on their direction of motion. This inherently requires knowledge of
their velocity, which precludes the usage of VHCs. 
Many authors have attempted to extend the theory of VHCs to enforce relations 
\(h(q,\dot{q}) = 0\) of the full state to account for this drawback. 
Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and has shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, the nonholonomic approach has been shown to be more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
This suggests that virtual nonholonomic constraints are more capable of
stabilizing specific energy levels while rejecting disturbances, all while
producing realistic biological motion. It is for this reason we choose to study
nonholonomic constraints in this thesis.

Unlike the theory of VHCs, there does not appear to be a standard definition of
virtual nonholonomic constraints: 
all the applications listed above use their own definition of a virtual
nonholonomic constraint, which makes it difficult to compare and analyze their
work.

This section will provide a new characterization of virtual nonholonomic
constraints in the Hamiltonian framework. 
The goal is to provide a consistent, rigorous foundation for
designing constraints on a general class of systems.

@@

If we define the error term  \(e = h(q,p)\), stabilizing \(\Gamma\) is
equivalent to solving for \(\tau(q,p)\) which drives \(e \rightarrow 0\) and
\(\dot{e} \rightarrow 0\).
Let us imagine that we have no further structure on the VNHC. 
Then \(\dot{e} = dh_q \dot{q} + dh_p \dot{p}\), where \(\tau\) appears inside
\(\dot{p}\). To solve for \(\tau\) explicitly, we must have that \(dh_p\) is
invertible for all \((q,p)\).
%which is the same as requiring that \(h(q,p)\) is strictly monotonic in \(p\).

This is a very restrictive condition, which many VNHCs will not satisfy. For
this reason, it is preferable to have the torque \(\tau\) appear after two 
derivatives of \(e\) so that there is more freedom in the types of constraints
one can use.
Mathematically, if each \(\tau_i\) appears only after two derivatives, one says 
that \(e\) is of \textit{relative degree} \(\{2,2,\ldots,2\}\). 
We thus define a special type of VNHC which satisfies this
property.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that a relation which uses only the unactuated conjugate of momentum
\(p_u\) cannot have \(\tau\) appearing after only one derivative. Of course,
they performed their research in Lagrangian form; 
for us to be able to use \(p_u\) in its Hamiltonian form, 
we will continue with the following assumption for the rest of the chapter.

@@

    That is, a regular VNHC is a function with domain 
    \(\mathcal{Q} \times \mathcal{P}_u\) which satisfies the rank condition.

@@

    This means that the entire matrix is given by

@@

        \Hess_q\left\{h\right\}\tpose -{} \\

@@

@@

@@

@@

    Moving on to \(\diff{}{t}\left\{dh_{p_u}\right\}\), a similar approach as

@@

@@

        \partial_q\partial_{p_u}h -{} \\

@@

    Define \(C_2 : \mathcal{Q} \times \mathcal{P} \rightarrow \R^{n(n-k)\times n}\) 

@@

    In matrix form, this becomes:
        & \frac{1}{2} \diff{}{t} \left\{ \pudmat \right\} ={} \\

@@

        \\

@@

        \ddot{e} &= \diff{}{t}\left\{dh_q\right\}\Minv(q)p + dh_q C_1(q,p)p -{}\\

@@

        dh_q\Minv(q)\nabla_qV(q) -{}\\

@@

     \left(\frac{1}{2}\pudmat + \nabla_{q_u}V(q)\right) +{} \\

@@

     \frac{1}{2}\pudmat + \nabla_{q_u}V(q)\right) -{}\\

@@

     dh_{p_u} \partial_{q_u}\partial_q V(q)\Minv(q)p +{} \\

@@

% TODO: Do we need the assumption that one can solve for qa linearly?
\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a ----- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated}. Let
    \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) 
    with constraint manifold \(\Gamma\). Suppose that on \(\Gamma\) one can
    solve linearly for \(q_a\) as a function of \((q_u,p_u)\).
    Then the closed-loop dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
            \dot{q}_u &= M(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}\right|_{\begin{array}{c}
                h(q,p_u) = 0 \\ 
                pa = g(q_u,p_u) \\
            \end{array}}
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        g(q,p_u) &:= \\
           &\left.\left(dh_q \Minv(q) \simpleB \right)\inv 
        \left(dh_{p_u} \nabla_{q_u}V(q) - dh_q \Minv(q)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
        \end{bmatrix} p_u\right)\right|_{h(q,p_u) = 0}
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    Setting \(e = h(q,p_u)\) and using the fact that 
    \(\nabla_{q_u}\Minv(q) = 0\), we find that
    \[
        \dot{e} = dh_q\Minv(q)p - dh_{p_u}\nabla_{q_u}V(q)
    \]
    Observe that
    \begin{align*}
        dh_q\Minv(q)p &= dh_q\Minv(q) \begin{bmatrix}p_u \\ p_a \end{bmatrix} \\
          &= dh_q\Minv(q) \begin{bmatrix}
              \Id{n-k} & \Zmat{(n-k)\times k} \\
              \Zmat{k \times(n-k)} & \Id{k} 
              \end{bmatrix} \begin{bmatrix} p_u \\ p_a \end{bmatrix} \\
        &= dh_q\Minv(q)\begin{bmatrix} \Id{n-k} \\ \Zmat{k \times
        (n-k)}\end{bmatrix} p_u + dh_q \Minv(q)\simpleB p_a
    \end{align*}
    On the constraint manifold, we have \(e = \dot{e} = 0\), which means
    \[
        dh_q\Minv(q)\simpleB p_a = dh_{p_u}\nabla_{q_u}V(q) -
            dh_q\Minv(q)\begin{bmatrix} 
            \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u
    \]
    Since \(h\) is regular and \(\nabla_{q_u}\Minv(q) = 0\), we have that 
    \[
        \rank{dh_q\Minv(q)\simpleB} = k
    \]
    Solving for \(p_a\) gives
    \[
        p_a(q,p_u) = \left(dh_q\Minv(q)\simpleB\right)\inv
        \left(dh_{p_u}\nabla_{q_u}V(q) - 
            dh_q\Minv(q)\begin{bmatrix} 
        \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u \right)
    \]
    This yields a function \(p_a(q,p_u)\). However, on \(\Gamma\) we have 
    \(h(q,p_u) = 0\) and can solve for \(q_a\) in terms of \((q_u,p_u)\);
    hence, we can solve for \(p_a = g(q_u,p_u)\). Since \(q_a\) and
    \(p_a\) can be computed directly from \((q_u,p_u)\), the dynamics on
    \(\Gamma\) are parameterized only by \((\dot{q}_u,\dot{p}_u)\).

>>>>

\section{Virtual Nonholonomic Constraints}
% Motivation for VNHCs, they are an extension of VHCs, etc.
Let us imagine a child on a swing who wants to reach the largest height
possible. 
To begin, the child pushes off the ground to imbue the swing with small oscillations.
What allows them to increase the amplitude of these oscillations is the
appropriate extension and retraction of their feet.
If a roboticist were creating a machine to replicate this behaviour, they might
design a robot whose legs extend and retract at specific time intervals. 
At first glance, this technique should work perfectly because the leg motion
would synchronize with the swinging frequency, thereby injecting energy as
quickly as is physically possible.

Unfortunately, a deeper analysis reveals the flaw with this design.
Most children are not counting out the time in their head; rather, they observe
their current position and velocity and adjust their legs as required. 
For example, many children have an adult pushing the swing, or perhaps
they are swinging on a windy day. In either case, they adjust their leg
motion accordingly when presented with these external disturbances, without
keeping track of time. 
Hence, the standard control technique of tracking a function of time (known as
\textit{trajectory tracking}) does not truly replicate human behaviour.
Even if the robot's legs perfectly track a specified trajectory, 
an external disturbance will desynchronize the leg motion
with the swing - thereby stopping the amplitude-increasing effects.

%As has been observed in the field of mobile robotics, a trajectory tracking
%controller which has lagged behind due to a disturbance will do whatever is
%required to ``catch up" to the trajectory. 
%This can cause major safety issues, which is why many mobile roboticists prefer to
%design controllers which drive a vehicle along a gemoetric path in space rather than a
%trajectory in time.

%This path tracking approach has been extended to biologically-inspired robotics
%in a method known as \textit{virtual holonomic constraints} (VHCs).
Rather than tracking a trajectory over time, a more human-like behaviour 
is to force the robot's legs track a function of the swing's state. 
One recent control method known as \textit{virtual holonomic constraints} (VHCs)
uses the actuators to enforce a relation \(h(q) = 0\) of the configuration
\cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like swimming robot
\cite{vhc_snake}.

The downside to VHCs is that they do not use the additional information imparted
by the velocity of a system.
For the child on a swing, whether they extend or retract their legs
depends on their direction of motion. 
This inherently requires knowledge of their current velocity, which precludes
the usage of VHCs. 
Many authors have attempted to extend the theory of VHCs to enforce relations
\(h(q,\dot{q}) = 0\) of the full state to account for this drawback. 
Since these relations use actuators to restrict both the configuration and
velocity of a system, they are called virtual \textit{nonholonomic} constraints.
This idea has been used for human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and has shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, this nonholonomic approach is more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
In particular, virtual nonholonomic constraints may be capable of injecting and
dissipating energy from a system in a robust manner, all while producing
realistic biological motion. This is what we aim to prove in this thesis.

Unlike the theory of VHCs, there does not appear to be a standard definition of
virtual nonholonomic constraints: 
all the applications listed above use their own definitions, which makes it
difficult to compare and generalize their work.

This section will provide a standard characterization of virtual nonholonomic
constraints using Hamiltonian mechanics. 
The goal is to provide a consistent, rigorous foundation for designing
constraints on a general class of systems.

@@

If we define the error term  \(e = h(q,p)\), stabilizing \(\Gamma\) is
equivalent to solving for the controller \(\tau(q,p)\) which drives 
\(e \rightarrow 0\) and \(\dot{e} \rightarrow 0\).
Supposing there is no further structure on the VNHC, 
then \(\dot{e} = dh_q \dot{q} + dh_p \dot{p}\). Since \(B(q) \tau\) appears inside
\(\dot{p}\), solving for \(\tau\) explicitly requires 
\(dh_p(q,p)B(q) \in \R^{k \times k}\) to be invertible for all \((q,p)\).
%which is the same as requiring that \(h(q,p)\) is strictly monotonic in \(p\).

This condition severely restricts the types of constraints one can use. 
It is preferable instead to have the torque \(\tau\) appear after two
derivatives of \(e\) to allow for more freedom in choosing a constraint.
Mathematically, if each \(\tau_i\) appears only after two derivatives, one says 
that \(e\) is of \textit{relative degree} \(\{2,2,\ldots,2\}\). 
We thus define a special type of VNHC which satisfies this
property.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that a relation which uses only the unactuated conjugate of momentum
cannot have \(\tau\) appearing after only one derivative. Of course,
they performed their research in Lagrangian form; 
for us to be able to use the Hamiltonian coordinate \(p_u\), 
we must continue with the following assumption for the rest of the chapter.

@@

    That is, a regular VNHC is a function 
    \(h : \mathcal{Q} \times \mathcal{P}_u \rightarrow \R^k\) which satisfies
    the rank condition.

@@

    This means that \(\diff{}{t}\left\{dh_q\right\}\) is given by
    \begin{align}

@@

        \Hess_q\left\{h\right\}\tpose -{} 
        \nonumber \\

@@

    \end{align}

@@

    \begin{equation}

@@

    \end{equation}

@@

    Moving on to \(\diff{}{t}\left\{dh_{p_u}\right\}\), a similar approach to

@@

    \begin{align}

@@

        \partial_q\partial_{p_u}h -{} 
        \nonumber \\

@@

    Defining \(C_2 : \mathcal{Q} \times \mathcal{P} \rightarrow \R^{n(n-k)\times n}\) 

@@

    \end{equation}
    allows us to collapse the derivative into:
    \begin{align}
        & \frac{1}{2} \diff{}{t} \left\{ \pudmat \right\} ={} 
        \nonumber \\

@@

        \nonumber \\

@@

    \begin{align}
        \ddot{e} &= \diff{}{t}\left\{dh_q\right\}\Minv(q)p + dh_q C_1(q,p)p -{}
        \nonumber \\

@@

        dh_q\Minv(q)\nabla_qV(q) -{} 
        \nonumber \\

@@

     \left(\frac{1}{2}\pudmat + \nabla_{q_u}V(q)\right) +{} 
     \nonumber \\

@@

     \frac{1}{2}\pudmat + \nabla_{q_u}V(q)\right) -{}
     \nonumber \\

@@

     dh_{p_u} \partial_{q_u}\partial_q V(q)\Minv(q)p +{} 
     \nonumber \\

@@

% TODO: Do we need the assumption that one can solve for qa linearly?
\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a ----- mechanical system satisfying Assumption
    \ref{assm:H-is-simply-actuated}. Let
    \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) 
    with constraint manifold \(\Gamma\). Suppose that on \(\Gamma\) one can
    solve linearly for \(q_a\) as a function of \((q_u,p_u)\).
    Then the closed-loop dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
                \dot{q}_u &= \begin{bmatrix}
                    \Id{(n-k)} & \Zmat{(n-k) \times k}
                \end{bmatrix}\Minv(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}{}\right|_{\begin{array}{c}
                h(q,p_u) = 0 \\ 
                p_a = g(q_u,p_u) \\
            \end{array}}
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        g(q,p_u) &:= \\
           &\left.\left(dh_q \Minv(q) \simpleB \right)\inv 
        \left(dh_{p_u} \nabla_{q_u}V(q) - dh_q \Minv(q)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
        \end{bmatrix} p_u\right)\right|_{h(q,p_u) = 0}
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    Setting \(e = h(q,p_u)\) and using the fact that 
    \(\nabla_{q_u}\Minv(q) = 0\), we find that
    \[
        \dot{e} = dh_q\Minv(q)p - dh_{p_u}\nabla_{q_u}V(q)
    \]
    Observe that
    \begin{align*}
        dh_q\Minv(q)p &= dh_q\Minv(q) \begin{bmatrix}p_u \\ p_a \end{bmatrix} \\
          &= dh_q\Minv(q) \begin{bmatrix}
              \Id{n-k} & \Zmat{(n-k)\times k} \\
              \Zmat{k \times(n-k)} & \Id{k} 
              \end{bmatrix} \begin{bmatrix} p_u \\ p_a \end{bmatrix} \\
        &= dh_q\Minv(q)\begin{bmatrix} \Id{n-k} \\ \Zmat{k \times
        (n-k)}\end{bmatrix} p_u + dh_q \Minv(q)\simpleB p_a
    \end{align*}
    On the constraint manifold, we have \(e = \dot{e} = 0\), which means
    \[
        dh_q\Minv(q)\simpleB p_a = dh_{p_u}\nabla_{q_u}V(q) -
            dh_q\Minv(q)\begin{bmatrix} 
            \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u
    \]
    Since \(h\) is regular and \(\nabla_{q_u}\Minv(q) = 0\), we have that 
    \[
        \rank{dh_q\Minv(q)\simpleB} = k
    \]
    Solving for \(p_a\) gives
    \[
        p_a(q,p_u) = \left(dh_q\Minv(q)\simpleB\right)\inv
        \left(dh_{p_u}\nabla_{q_u}V(q) - 
            dh_q\Minv(q)\begin{bmatrix} 
        \Id{n-k} \\ \Zmat{k \times (n-k)}\end{bmatrix} p_u \right)
    \]
    This yields a function \(p_a(q,p_u)\). However, on \(\Gamma\) we have 
    \(h(q,p_u) = 0\) and can solve for \(q_a\) in terms of \((q_u,p_u)\);
    hence, we can solve for \(p_a = g(q_u,p_u)\). Since \(q_a\) and
    \(p_a\) can be computed directly from \((q_u,p_u)\), the dynamics on
    \(\Gamma\) are parameterized only by \((\dot{q}_u,\dot{p}_u)\).
